Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-12 jss/EpiCompare-paper/EpiCompare-paper.Rmd
updates to section 3 (2/3rds of it anyway).

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with EpiCompare"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhances current infectious disease analysis pipelines and encourages comparisons across models and epidemics. A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands. \pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to make adoption of the package easier and analysis faster. This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{afterpage} 
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{rotating}
---

@@

\ben{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Fig. \ref{fig:pipeline2}, we illustrate how our package fits into the data analysis pipeline introduced in Fig. \ref{fig:pipeline}.  All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}. Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.

@@

@@

The first step of most data analysis is cleaning up the data to be explored. There are multiple ways to collect the epidemiological data. Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information will be collected, whereas other data collections will focus on aggregate counts of individuals in each epidemic state.  In fact, usually only the number of new infections at each time step is observed and the compartment totals are then imputed from that along with other information about the disease and the population of interest.  We focus on understanding the overall impact of an outbreak at the aggregate/population level.  In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g. start of infection, start of recovery, etc). This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations). Our tools allow the user to easily group agents and define new subpopulations to explore. This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g. age, sex) and many state based models provide for subpopulation-based states in their analysis \citep{rvachev1985,anderson1992,worby2015}.

We provide a "tidyverse"-styled function, `agents_to_aggregate()` to transform agent information into aggregate state information.  As a "tidy" function, our function `agents_to_aggregate()` allows the user to identify and analyze subpopulations by first executing a `group_by()` from \pkg{dplyr}. The function `agents_to_aggregate` is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates. Our function, `agents_to_aggregate()` does not require agents to pass through every state listed and agents are allowed to start in any state. This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g. hours or minutes). The function `agents_to_aggregate()` returns the total number of individuals in each state at each time point.  We view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.

@@

With raw data,  exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from `agents_to_aggregate()` to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model). Our `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind `agents_to_aggregate()` with \pkg{ggplot2}'s `geom_path()` and \pkg{ggtern}'s `coord_tern()` to visualize any number of groups epidemic trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}. We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).

**Model Fitting and Simulations**

Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next.  These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.  The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.  This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

>>>>

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with EpiCompare"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhances current infectious disease analysis pipelines and encourages comparisons across models and epidemics. A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands. \pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to make adoption of the package easier and analysis faster. This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{afterpage}
  \usepackage[normalem]{ulem} 
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{rotating}
---

@@

@@

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Fig. \ref{fig:pipeline2}, we illustrate how our \ben{\sout{package fits} package's functions fit} into the data analysis pipeline introduced in Fig. \ref{fig:pipeline}.  All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}. Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.

@@

The first step of most data analysis is cleaning up the data \ben{so it can \sout{to}} be explored. There are multiple ways to collect the epidemiological data. Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections will focus on aggregate counts of individuals in each epidemic state.  In fact, usually only the number of new infections at each time step is observed and the compartment totals \ben{(amounts of individuals in each state)} are then imputed from that along with other information about the disease and the population of interest.  \ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows examination of overall trends of an epidemic more easily}.

\ben{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }`agents_to_aggregate`\ben{. The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}

\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner to a point in which they can examine their epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`. This function transforms data that contains information about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc) to an aggregate view of how many individuals where in a state at a given time. There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. Papers like \citet{rvachev1985,anderson1992,worby2015} also develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}. This function allows the user to easily group agents into new subpopulations by using the \pkg{tidyverse} paradigm to first `group_by()`^[The `group_by` function is from the \pkg{dplyr} package.] the subpopulation defining features and then aggregate (using `agents_to_aggregate`) conditional on these subpopulations.

\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is also flexible to deal with a wide range of information about each individual. It can account for infinitely many states, allowing the practioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more. Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall number of individuals in the population). This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data. For these reasons, we view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.

\ben{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g. start of infection, start of recovery, etc). This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations). Our tools allow the user to easily group agents and define new subpopulations to explore. This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g. age, sex) and many state based models provide for subpopulation-based states in their analysis}} \citep{rvachev1985,anderson1992,worby2015}.

\ben{\sout{We provide a "tidyverse"-styled function, }`agents_to_aggregate()`\ben{\sout{ to transform agent information into aggregate state information.  As a "tidy" function, our function }}`agents_to_aggregate()` \ben{\sout{allows the user to identify and analyze subpopulations by first executing a }}`group_by()` \ben{\sout{{ from \pkg{dplyr}. The function }}`agents_to_aggregate`\ben{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates. Our function,}} `agents_to_aggregate()`\ben{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state. This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g. hours or minutes). The function }}`agents_to_aggregate()`\ben{\sout{ returns the total number of individuals in each state at each time point.  We view }}`agents_to_aggregate()`\ben{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}

@@

\ben{[Ben says: I'm went back and thought that the version before this version was clearer. Wanna take a look and discuss? Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}

\ben{[Ben says: this is actually an older version:]} With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting. An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model). Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).

\ben{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}

\ben{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from}} `agents_to_aggregate()` \ben{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model). Our}} `geom_aggregate()` \ben{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind}} `agents_to_aggregate()`  \ben{\sout{with \pkg{ggplot2}'s} `geom_path()`  \ben{\sout{and \pkg{ggtern}'s}} `coord_tern()`  \ben{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}} \citep{Wickham2016, Hamilton2018}.\ben{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}

**Model Fitting and Simulations**
\ben{[Ben says: see footnote 3 about the introduction to this paragraph.]}

Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next^[The original sentence seemed clearer to me (it may even be a run-on sentence now). I'm wondering if you didn't like the original sentence relative to the "truth-ful-ness" of the statement? Here's the original sentence slightly edited: "Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents".].  These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.  The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.  This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

### 2021-03-12 jss/EpiCompare-paper/EpiCompare-paper.tex
updates to section 3 (2/3rds of it anyway).

\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage}

@@

\textcolor{violet}{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our package fits into the data
analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible, and
we focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight both approaches and how to leverage \pkg{EpiCompare} in the
subsections below.

@@

@@

The first step of most data analysis is cleaning up the data to be
explored. There are multiple ways to collect the epidemiological data.
Sometimes individual records with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population information will be
collected, whereas other data collections will focus on aggregate counts
of individuals in each epidemic state. In fact, usually only the number
of new infections at each time step is observed and the compartment
totals are then imputed from that along with other information about the
disease and the population of interest. We focus on understanding the
overall impact of an outbreak at the aggregate/population level. In
\pkg{EpiCompare}, we provide a function to transform information about
each individual/agent's initial time of entry into each state
(e.g.~start of infection, start of recovery, etc). This transformation
between \textit{agent} information to \textit{aggregate} information is
useful for seeing the overall trend of the epidemic (and how it impacts
different subpopulations). Our tools allow the user to easily group
agents and define new subpopulations to explore. This is important as
case studies highlight the usefulness of identifying differing
subpopulations (e.g.~age, sex) and many state based models provide for
subpopulation-based states in their analysis
\citep{rvachev1985,anderson1992,worby2015}.

We provide a ``tidyverse''-styled function,
\texttt{agents\_to\_aggregate()} to transform agent information into
aggregate state information. As a ``tidy'' function, our function
\texttt{agents\_to\_aggregate()} allows the user to identify and analyze
subpopulations by first executing a \texttt{group\_by()} from
\pkg{dplyr}. The function \texttt{agents\_to\_aggregate} is flexible in
that there is no limit on the amount of permitted epidemiological states
so users can incorporate information from compartments such as
``Exposed'', ``iMmune'', or ``Hospitalized'' groups, for instance, and
also permits indicators for death/exit and birth/entry dates. Our
function, \texttt{agents\_to\_aggregate()} does not require agents to
pass through every state listed and agents are allowed to start in any
state. This function is constrained to integer time steps (for example
days), but transformations (linear or otherwise) of the time columns can
account for finer grained time steps (e.g.~hours or minutes). The
function \texttt{agents\_to\_aggregate()} returns the total number of
individuals in each state at each time point. We view
\texttt{agents\_to\_aggregate()} as powerful tool in the step of
pre-processing data, one that quickly and simply transforms individual
level information into an aggregate view that can be processed and
analyzed in a simpler manner.

@@

With raw data, exploratory analysis frequently means figuring out good
combinations of visualizations, numerical summaries, and groupings. An
expert coder can start from \texttt{agents\_to\_aggregate()} to
successfully accomplish EDA in many ways, but we have also developed
tools to allow a novice coder to rapidly explore data, as long as there
three unique epidemiological states (like the SIR model). Our
\texttt{geom\_aggregate()} provides a rapid way to explore different
subpopulations' experiences of the epidemic in the ternary lens. It
combines the ideas behind \texttt{agents\_to\_aggregate()} with
\pkg{ggplot2}'s \texttt{geom\_path()} and \pkg{ggtern}'s
\texttt{coord\_tern()} to visualize any number of groups epidemic
trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}. We developed
these out-of-the-box visualization tools for SIR models because (1) SIR
models are some of the most common and basic epidemic state-based models
and (2) our simplex representation of these epidemics emphasizes a
``time-invarance'' representation of the data (see Section
\ref{sec:time-invariant}).

\textbf{Model Fitting and Simulations}

Although this package does not focus on estimating a model for the data,
we do provide some flexible functions for simulation of basic
discrete-time epidemic-state models with Bernoulli/Multinomial
transitions which can be used to (noisily) approximate ODEs which
describe transitions from one state to the next. These functions can be
naturally combined with \texttt{agents\_to\_aggregate()} to proceed in
the pipeline. The function \texttt{simulate\_SIR\_agents()} simulates an
SIR epidemic with user inputs for the number of simulations, the initial
number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified transition matrix and other epidemic parameters
to allow the user to create simulations of an outbreak for \textit{any}
number of states and any number of transitions among them. This allows
for users to explore the space of models in an intuitive way without
getting bogged down by too much mathematical detail. For consistency, we
have made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.

>>>>

\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage} \usepackage[normalem]{ulem}

@@

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our
\textcolor{violet}{\sout{package fits} package's functions fit} into the
data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible, and
we focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight both approaches and how to leverage \pkg{EpiCompare} in the
subsections below.

@@

The first step of most data analysis is cleaning up the data
\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
ways to collect the epidemiological data. Sometimes individual records
\textcolor{violet}{are collected}, with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population
information\textcolor{violet}{.\sout{will be collected, whereas}}
\textcolor{violet}{O}ther data collections will focus on aggregate
counts of individuals in each epidemic state. In fact, usually only the
number of new infections at each time step is observed and the
compartment totals
\textcolor{violet}{(amounts of individuals in each state)} are then
imputed from that along with other information about the disease and the
population of interest. \textcolor{violet}{In \pkg{EpiCompare}, w}e
focus on understanding the overall impact of an outbreak at the
aggregate/population level,
\textcolor{violet}{which allows examination of overall trends of an epidemic more easily}.

\textcolor{violet}{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }\texttt{agents\_to\_aggregate}\textcolor{violet}{. The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}

\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
help the practitioner to a point in which they can examine their
epidemics from an aggregate/ population lens, we provide a function
called \texttt{agents\_to\_aggregate}. This function transforms data
that contains information about individual/agents' initial entry into
each state (e.g.~start of infection, start of recovery, etc) to an
aggregate view of how many individuals where in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. Papers like
\citet{rvachev1985,anderson1992,worby2015} also develop state-based
models that account for subpopulations in their
models\footnote{\textcolor{violet}{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
This function allows the user to easily group agents into new
subpopulations by using the \pkg{tidyverse} paradigm to first
\texttt{group\_by()}\footnote{The \texttt{group\_by} function is from
  the \pkg{dplyr} package.} the subpopulation defining features and then
aggregate (using \texttt{agents\_to\_aggregate}) conditional on these
subpopulations.

\textcolor{violet}{[Ben says: this whole paragraph is new]} The
\texttt{agents\_to\_aggregate} function is also flexible to deal with a
wide range of information about each individual. It can account for
infinitely many states, allowing the practioner to aggregate information
relative to the standard ``Suspectible'', ``Infectious'', and
``Recovered'' states (CITE) or also add states like ``Exposed'',
``iMmune'', ``Hospitalized'' and more. Additionally,
\texttt{agents\_to\_aggregate} also permits indicators for death/exit
and birth/entry dates (that change the overall number of individuals in
the population). This function is the highest tested function in this
package, and can deal with many nuances that might come-about in the
individual data. For these reasons, we view
\texttt{agents\_to\_aggregate()} as powerful tool in the step of
pre-processing data, one that quickly and simply transforms individual
level information into an aggregate view that can be processed and
analyzed in a simpler manner.

\textcolor{violet}{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g. start of infection, start of recovery, etc). This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations). Our tools allow the user to easily group agents and define new subpopulations to explore. This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g. age, sex) and many state based models provide for subpopulation-based states in their analysis}}
\citep{rvachev1985,anderson1992,worby2015}.

\textbackslash ben\{\sout{We provide a "tidyverse"-styled function, }\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ to transform agent information into aggregate state information.  As a "tidy" function, our function }}\texttt{agents\_to\_aggregate()}
\textcolor{violet}{\sout{allows the user to identify and analyze subpopulations by first executing a }}\texttt{group\_by()}
\textbackslash ben\{\sout{{ from \pkg{dplyr}. The function }}\texttt{agents\_to\_aggregate}\textcolor{violet}{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates. Our function,}}
\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state. This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g. hours or minutes). The function }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ returns the total number of individuals in each state at each time point.  We view }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}

@@

\textcolor{violet}{[Ben says: I'm went back and thought that the version before this version was clearer. Wanna take a look and discuss? Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}

\textcolor{violet}{[Ben says: this is actually an older version:]} With
raw data, getting to know our data now-a-days very frequently means
figuring out good combinations of visualizations, numerical summaries
and subsetting. An expert coder can start from
\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
we've also developed tools to rapidly explore your data if your have a 3
unique states models (like the SIR model). Our \texttt{geom\_aggregate}
provides a rapid way to explore different subpopulations experience of
the epidemic, and combines the ideas behind agents\_to\_aggregate for
the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
visualize any number of groups epidemic trajectory in 3d simplex space
using \texttt{ggplot2} and \texttt{ggtern}
\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
were developed because (1) SIR models are some of the most common and
basic epidemic state-based models and (2) our simplex representation of
these epidemics emphasizes a ``time-invarance'' representation of the
data (for a refresher see Section \ref{sec:time-invariant}).

\textcolor{violet}{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}

\textcolor{violet}{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from}}
\texttt{agents\_to\_aggregate()}
\textcolor{violet}{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model). Our}}
\texttt{geom\_aggregate()}
\textcolor{violet}{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind}}
\texttt{agents\_to\_aggregate()}
\textbackslash ben\{\sout{with \pkg{ggplot2}'s} \texttt{geom\_path()}
\textcolor{violet}{\sout{and \pkg{ggtern}'s}} \texttt{coord\_tern()}
\textcolor{violet}{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}}
\citep{Wickham2016, Hamilton2018}.\textcolor{violet}{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}

\textbf{Model Fitting and Simulations}
\textcolor{violet}{[Ben says: see footnote 3 about the introduction to this paragraph.]}

Although this package does not focus on estimating a model for the data,
we do provide some flexible functions for simulation of basic
discrete-time epidemic-state models with Bernoulli/Multinomial
transitions which can be used to (noisily) approximate ODEs which
describe transitions from one state to the next\footnote{The original
  sentence seemed clearer to me (it may even be a run-on sentence now).
  I'm wondering if you didn't like the original sentence relative to the
  ``truth-ful-ness'' of the statement? Here's the original sentence
  slightly edited: ``Although this package does not focus on estimating
  a model for the data, we do provide some power functions for
  simulation of basic state models with Bernoulli approximation of ODE
  state models encoded in simulate\_agents and simulate\_SIR\_agents''.}.
These functions can be naturally combined with
\texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
function \texttt{simulate\_SIR\_agents()} simulates an SIR epidemic with
user inputs for the number of simulations, the initial number in each
state, the infection and recovery parameters \((\beta, \gamma\)), and
the total number of discrete time steps. This function allows for easy
access to SIR model analysis and comparison. Beyond SIR models, the
function \texttt{simulate\_agents()} takes as input a user-specified
transition matrix and other epidemic parameters to allow the user to
create simulations of an outbreak for \textit{any} number of states and
any number of transitions among them. This allows for users to explore
the space of models in an intuitive way without getting bogged down by
too much mathematical detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.
